---
title: "SIP-mav3 for Malaria"
subtitle: "A Generalized SIP Model for Malaria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIP-mav3 for Malaria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(viridisLite)
library(ramp.library)
library(deSolve)
#devtools::load_all()
```


```{r, echo=FALSE}
mav3_opts = list()
mav3_opts$mod_b = "im" 
mav3_opts$b_opts = list(b0 = .55, Nb=20) 
mav3_opts$mod_severe_h = "protect" 
mav3_opts$severe_h_opts = list(p0 = .55, Np=1) 
mav3_opts$mod_severe_x = "protect" 
mav3_opts$severe_x_opts = list(rate0 = 1/300, Np=1) 
mav3_opts$mod_moderate_h = "protect" 
mav3_opts$moderate_h_opts = list(p0 = 1, Np=20) 
mav3_opts$mod_moderate_x = "protect" 
mav3_opts$moderate_x_opts = list(rate0 = 1/300, Np=20) 
mav3_opts$mod_mild_h = "protect" 
mav3_opts$mild_h_opts = list(p0 = 1, Np=100) 
mav3_opts$mod_mild_x = "protect" 
mav3_opts$mild_x_opts = list(rate0 = 1/20, Np=100) 
mav3_opts$mod_c = "dim" 
mav3_opts$c_opts = list(c0=.8, pda=80, Nc=4) 

mod <- xds_setup_cohort(Xname = "SIPmav3", Xopts = mav3_opts, age_par = makepar_F_type2())
mod <- setup_exposure_nb(mod, .1)
```

```{r}

mod$EIRpar$eir = .2/365
mod0 <- xds_solve_cohort(mod, A=10, da=1)
XH0 <- get_XH(mod0)

mod$EIRpar$eir = 1/365
mod1 <- xds_solve_cohort(mod, A=10, da=1)
XH1 <- get_XH(mod1)

mod$EIRpar$eir = 5/365
mod5 <- xds_solve_cohort(mod, A=10, da=1)
XH5 <- get_XH(mod5)

mod$EIRpar$eir = 25/365
mod25 <- xds_solve_cohort(mod, A=10, da=1)
XH25 <- get_XH(mod25)
```

## Exposure

### Exposure $\times$ age 

+ Set up by `xds_setup_cohort(..., age_par = make_function(makepar_F_type2()))` 

To model biting by age, we use a simple function:


```{r, fig.height=5, fig.width=8}
omega = make_function(makepar_F_type2())
aa = seq(0, 7300, by=10)
plot(aa/365, omega(aa), type = "l", ylim = c(0, 1.5), xlab = "Age (Years)", ylab = expression(omega(a)))
segments(0,1, 20, 1, lty =2)
```


### Infections per Infectious Bite

We let 

$$b = F_b(v_h) = b_0 exp(-v_h/N_b)$$ 

At three levels of exposure, this is what $F_b(a)$ looks like: 


```{r, fig.height=5, fig.width=8}
clrs <- turbo(7)
age <- XH1$time

b_a_0 = with(mod$Xpar[[1]]$b_mod, F_infect_im(XH0$vh, b0, Nb))
b_a_1 = with(mod$Xpar[[1]]$b_mod, F_infect_im(XH1$vh, b0, Nb))
b_a_3 = with(mod$Xpar[[1]]$b_mod, F_infect_im(XH5$vh, b0, Nb))
b_a_10 = with(mod$Xpar[[1]]$b_mod, F_infect_im(XH25$vh, b0, Nb))
plot(age/365, b_a_1, type="l", xlab = "Age (Years)", ylab = expression(F[b](a)), ylim = c(0,1), col = clrs[2])
lines(age/365, b_a_3, col = clrs[4])
lines(age/365, b_a_10, col = clrs[6])
lines(age/365, b_a_0, col = clrs[1])
segments(0, 0.55, 20, 0.55, lty =2)
```

### EIR to FoI 

At three levels of exposure, we plot $F_b(v_h(a)) \omega(a)$: 

```{r, fig.height=5, fig.width=8}
w_a <- omega(age)
plot(age/365, 0.55*w_a, type="l", xlab = "Age (Years)", ylab=expression(list(omega(a), omega(a)*b(a))), lwd=2)
lines(age/365, b_a_0*w_a, col = clrs[1])
lines(age/365, b_a_1*w_a, col = clrs[2])
lines(age/365, b_a_3*w_a, col = clrs[4])
lines(age/365, b_a_10*w_a, col = clrs[6])
```

### Environmental Heterogeneity

We set up a negative binomial model, with a size parameter $0.1.$

```
mod <- setup_exposure_nb(mod, .1)
```

### The FoI

Since the model outputs cumulative exposure, $v_h,$ we can diff it to get the realized FoI. The daily FoI with respect to age looks like this:

```{r, fig.height=5, fig.width=8}
a1 = age[-1]
foi0 <- diff(XH0$vh)
foi1 <- diff(XH1$vh)
foi5 <- diff(XH5$vh)
foi25 <- diff(XH25$vh)
plot(a1/365, foi25, type="l", xlab = "Age (Years)", ylab=expression(list(omega(a), omega(a)*b(a))), col = clrs[6], ylim = c(0, 0.014))
lines(a1/365, foi1, col = clrs[2])
lines(a1/365, foi5, col = clrs[4])
lines(a1/365, foi0, col = clrs[1])
```

Obviously, the effect of immunity is to homogenize exposure: We plot the three ratios: 

```{r, fig.height=5, fig.width=8}
plot(a1/365, foi25/foi1, type="l", xlab = "Age (Years)", ylab=expression(list(omega(a), omega(a)*b(a))), col = clrs[7], ylim = c(0,25))
lines(a1/365, foi25/foi5, type="l", col = clrs[5])
lines(a1/365, foi5/foi1, type="l", col = clrs[3])
lines(a1/365, foi1/foi0, type="l", col = clrs[1])
segments(0,1,10,1, col = grey(0.5))
#segments(0,3,10,3, col = grey(0.5))
#segments(0,2,10,2, col = grey(0.5))
```

## Malaria Prevalence 

I have to jigger some things to get prevalence by age to compute correctly. 

## Malaria Incidence

### Severe Malaria 

```{r, fig.height=5, fig.width=8}
severe0 = diff(XH0$severe)
severe1 = diff(XH1$severe)
severe5 = diff(XH5$severe)
severe25 = diff(XH25$severe)
plot(XH1$time[-1]/365, severe25*365, type = "l", col = clrs[6], ylab = "Severe Malaria, per Year", xlab = "Age (in Years)")
lines(XH1$time[-1]/365, severe5*365, col = clrs[4])
lines(XH1$time[-1]/365, severe1*365, col = clrs[2]) 
lines(XH0$time[-1]/365, severe0*365, col = clrs[1]) 

```

### Moderate Malaria 

```{r, fig.height=5, fig.width=8}
plot(XH1$time[-1]/365, diff(XH25$moderate)*365, type = "l", col = clrs[6], ylab = "moderate Malaria, per Year", xlab = "Age (in Years)", ylim = c(0,4))
lines(XH1$time[-1]/365, diff(XH5$moderate)*365, col = clrs[4])
lines(XH1$time[-1]/365, diff(XH1$moderate)*365, col = clrs[2]) 
lines(XH1$time[-1]/365, diff(XH0$moderate)*365, col = clrs[1]) 
```

### Mild Malaria 

```{r, fig.height=5, fig.width=8}
plot(XH1$time[-1]/365, diff(XH25$mild)*365, type = "l", col = clrs[6], ylab = "mild Malaria, per Year", xlab = "Age (in Years)", ylim = c(0,15))
lines(XH1$time[-1]/365, diff(XH5$mild)*365, col = clrs[4])
lines(XH1$time[-1]/365, diff(XH1$mild)*365, col = clrs[2]) 
lines(XH1$time[-1]/365, diff(XH0$mild)*365, col = clrs[1]) 
```

## Treatment 

```{r, fig.height=5, fig.width=8}
plot(XH1$time[-1]/365, diff(XH25$treat)*365, type = "l", col = clrs[6], ylab = "Yearly Treatment Rate", xlab = "Age (in Years)", ylim = c(0, 2000))
lines(XH1$time[-1]/365, diff(XH5$treat)*365, col = clrs[4])
lines(XH1$time[-1]/365, diff(XH1$treat)*365, col = clrs[2]) 
lines(XH1$time[-1]/365, diff(XH0$treat)*365, col = clrs[1]) 
```
