---
title: "SIP: an **XH** Module"
subtitle: "Susceptible, Infected, and post-treatment Protection" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIP: an **XH** Module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIP (Susceptible-Infected-Protected) human model model implements a basic model for human infection with malaria that includes an extra state for post-treatment chemoprotection, a short period when a person is protected from being infected. 

It is a reasonable first complication of the [`ramp.xds`::SIS human model](https://dd-harp.github.io/ramp.xds/articles/human_sis.html). This requires new parameters: $\rho$, the probability a new infection is treated; $\sigma$ is treatment of malaria infections; $\xi$ is background treatment; and $\eta$ the duration of chemoprophylaxis following treatment. $X$ remains a column vector giving the infectiuos density of infectious individuals in each strata, and $P$ the number of treated and protected individuals.

# Differential Equations

The equations are formulated around the FoI, $h$. Under the default model, we get the relationship $h = b E$, where $E$ is the daily EIR: 

$$
\frac{dH}{dt}  = F_B(t, H) - D \cdot H
$$


$$
\frac{dI}{dt}  = h (1-\rho) (H-I-P) - (r+\xi+\sigma)I + D \cdot I
$$

$$
\frac{dP}{dt} = h \rho (H-I-P) + \xi(H-P) + \sigma I - \eta P + D \cdot P
$$
In the model $$X = c I/ H.$$

# Equilibrium solutions

We can compute the steady states as a function of the FoI, $h$: 

$$
\bar I = H \frac{h  \eta (1-\rho)}{(h+r+\xi+\sigma)(\eta+\xi) +h((r-\eta)\rho+\sigma)}
$$
and 

$$
\bar P = H\frac{\xi(h+r+\xi+\sigma) + h(r\rho+\sigma)}{(h+r+\xi+\sigma)(\eta+\xi) +h((r-\eta)\rho+\sigma)}
$$
and 
$$\bar S = H - \bar I - \bar P$$ 

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(ramp.library)
library(deSolve)
library(data.table)
library(ggplot2)
```


```{r, echo=FALSE} 
#devtools::load_all()
```

Here we run a simple example with 3 population strata at equilibrium. We use `ramp.xds::make_parameters_X_SIP_xde` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time).

## The Long Way

```{r}
nStrata <- 3
H <- c(100, 500, 250)
nPatches <- 3
residence <- 1:3 
params <- make_xds_object_template("ode", "human", nPatches, 1, residence) 
```


```{r}
b <- 0.5
c <- 0.15
r <- 1/200
eta <- c(1/30, 1/40, 1/35)
rho <- c(0.05, 0.1, 0.15)
xi <- rep(0, 3)
Xo = list(b=b,c=c,r=r,eta=eta,rho=rho,xi=xi)
params = setup_XH_obj("SIP", params, 1, Xo) 
```

```{r}
eir <- c(1,2,3)/365
steady_state_X(eir*b, H, params) -> ss
params = setup_XH_inits(params, H, 1, ss)
```

```{r}
Xo <- c(Xo, ss) 
```


```{r}
MYo = list(
  Z = eir*H, f=1, q=1
)
```

```{r}
params = setup_MY_obj("trivial", params, 1, MYo)
params = setup_MY_inits(params, 1)
params = setup_L_obj("trivial", params, 1)
params = setup_L_inits(params, 1)
```


```{r}
params = make_indices(params)
```

```{r}
steady_state_X(eir*b, H, params, 1)
```


```{r}
y0 <- as.vector(unlist(get_inits(params)))
```

```{r} 
out <- deSolve::ode(y = y0, times = c(0, 730), xde_derivatives, parms= params, method = 'lsoda') 
get_XH_vars(out[2,-1], params, 1)
```
 
```{r, out.width = "100%"}
colnames(out)[params$XH_obj[[1]]$ix$H_ix+1] <- paste0('H_', 1:params$nStrata)
colnames(out)[params$XH_obj[[1]]$ix$I_ix+1] <- paste0('I_', 1:params$nStrata)
colnames(out)[params$XH_obj[[1]]$ix$P_ix+1] <- paste0('P_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  facet_wrap(. ~ Component, scales = 'free') +
  theme_bw()
```

## Using Setup


```{r}
xds_setup_human(Xname="SIP", nPatches=3, residence = 1:3, HPop=H, XHoptions= Xo, MYoptions = MYo) -> test_SIP_xde
```

```{r}
steady_state_X(b*eir, H, test_SIP_xde, 1) -> out1
out1 <- unlist(out1)
```

```{r}
xds_solve(test_SIP_xde, 365, 365)$outputs$last_y -> out2
approx_equal(out2,out1) 
```
