---
title: "Developing MY Modules" 
subtitle: "Models for Adult Mosquito Ecology and Infection Dynamics" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developing MY Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This vignette discusses how to add a new **MY** module to [`ramp.library`](https://github.com/dd-harp/ramp.library). As background, 
we strongly recommend reading the **`ramp.xds`** vignette about the [**MY** Interface.](https://dd-harp.github.io/ramp.xds/articles/MY-interface.html)

Each **MY** module implements a unique model family describing human/host infection dynamics and immunity that can be used by  [`ramp.xds`](https://github.com/dd-harp/ramp.xds). Before writing a new module, please check and see if the model family is already part of the SimBA library. 

If you need help, please contact [Professor David L Smith](mailto:smitdave@gmail.com)

*** 

We have written this vignette to produce a new template file using `knitr::purl.` The template can be modified to build a new **MY** module. 

Do this:

1. Download the markdown file that created this document ["X-How_To_Add_Model.Rmd."](https://github.com/dd-harp/ramp.library/blob/main/vignettes/X-How_To_Add_Model.Rmd)

2. Set your working directory to the location of the downloaded copy of "X-How_To_Add_Model.Rmd"

3. Execute these commands:

```{r, purl=F, eval=F}
library(knitr)
knitr::purl("X-How_To_Add_Model.Rmd", 
            output = "MYmodule-newMYname.R", 
            documentation = 0, 
            quiet=TRUE)
```

4. You should see a new file called `MYmodule-newMYname.R` in the working directory. You can now move and rename it. 

5. Open the file and do a global search for `newMYname` and replace it with the a short string that will define the model as its `Xname.`

6. Edit the functions following the directions below.  

## Style Guide 

We recommend that new models conform to a set of style guidelines:

+ All the names of the variables by name using the indices attached to `MY_obj` 
+ All documentation is written in `roxygen2`

+ Functions should use `with(*,{...})` to make the code easier to read. 

+ A vignette should accompany each model, including citations to the relevant publications, and proposed tests of the code.  

+ The functions should easy to relate to the equations defining a model, following notation defined in an accompanying vignette.

## Dynamics 


### dMYdt 

The sections that need to be modified are marked `# Change this`

+ Replace `M` with the your first variable (*e.g.* mosquito population density), and replace `F_M` with the formula that computes the derivatives. Do the same with `Y` and add lines for each variable.  

+ Modify `derivs = c(dM, dY, ...)` to return the LHS of the derivative equations you just wrote. 

```{r}
#' @title Compute the derivatives for parasite infection dynamics in human population strata 
#' @description Implements [dMYdt] for the newMYname model
#' @inheritParams ramp.xds::dMYdt
#' @return a [numeric] vector
#' @export
dMYdt.newMYname <- function(t, y, xds_obj, i) {

  # do not change this
  Lambda <- xds_obj$terms$Lambda[[i]]
  kappa <- xds_obj$terms$kappa[[i]]

  # do not change this
  # attach the variables by name 
  with(get_MY_vars(y, xds_obj, i),{
    # do not change this
    # attach the parameters by name 
    with(xds_obj$MY_obj[[i]], {
      # Do not change this 
      dM <- F_M(Lambda, M) 
      dY <- F_Y(kappa, M, Y) 
      ... 
      
      # Change this 
      derivs = c(dM, dY, ...) 
      
      # return the derivatives 
      return(derivs)
    })
  })
}
```

### Update_MYt 

```{r}
#' @title Compute the derivatives for parasite infection dynamics in human population strata 
#' @description Implements [UpdateMYt] for the newMYname model
#' @inheritParams ramp.xds::UpdateMYt
#' @return a [numeric] vector
#' @export
Update_MYt.newMYname<- function(t, y, xds_obj, s) {
  Lambda <- xds_obj$terms$Lambda[[i]]
  kappa <- xds_obj$terms$kappa[[i]]

   # do not change this
  # attach the variables by name 
  with(get_MY_vars(y, xds_obj, i),{
    # do not change this
    # attach the parameters by name 
    with(xds_obj$MY_obj[[i]], {
      # Change this
      Mt <- M + Lambda + F_M(M) 
      Yt <- Y + F_Y(kappa, M, Y) 
      ... 
      
      # Change this 
      states = c(Mt, Yt, ...) 
      
      # return the derivatives 
      return(states)
    })
  })
}
```


## MY Model Object 

### make_MY_obj

+ In the following, replace `p1` with the first parameter, `p2` with the second, and so on.

+ Unless you know what you're doing, leave the "Ports for human / host demography" unchanged. 

+ If you don't want to add ports for mass treatment, delete the line with `mda` and `msat` and leave `mda=FALSE` and `msat=FALSE` in  `skill_set_MY.newMYname.` If you do want your module to have that functionality, add the effects of mass treatment to the parameters that define state transitions in `dMYdt` or `Update_MYt` and set `mda=TRUE` and `msat=TRUE` in the skill set. 

```{r}
#' @title Make parameters for newMYname human model, with defaults
#' @param nPatches is the number of population strata
#' @param options a [list] that could overwrite defaults
#' @param p1 the first parameter 
#' @param p2 the second parameter 
#' @param p3 the third parameter 
#' @return a [list]
#' @export
make_MY_obj_newMYname = function(nPatches, options=list(),
                         p1=1, p2=2, p3=3){
  with(options,{
    MY_obj = list()
    class(MY_obj) <- c("newMYname")

    # Change this
    MY_obj$p1 = checkIt(p1, nPatches)
    MY_obj$p2 = checkIt(p2, nPatches)
    MY_obj$p3 = checkIt(p3, nPatches)

    # Don't change this
     MY_obj <- setup_K_obj(nPatches, MY_obj) 
    
    return(MY_obj)
  })}
```


### setup_MY_obj  

If you did a global search and replace on `newMYname,` this function will not need to be modified

```{r}
#' @title Setup MY_obj.newMYname
#' @description Implements [setup_MY_obj] for the newMYname model
#' @inheritParams ramp.xds::setup_MY_obj
#' @return a [list] vector
#' @export
setup_MY_obj.newMYname = function(newMYname, xds_obj, i, options=list()){
  xds_obj$MY_obj[[i]] = make_MY_obj_newMYname(xds_obj$nPatches, options)
  return(xds_obj)
}
```

## Parameters 

## Variables 

### setup_MY_ix  

In the following, each variable in your model gets its own pair of lines. Replace `M` in `M_ix` with the first variable name, `Y` in `Y_ix` with the second, and add lines for each variable by copying and pasting and editing the pair of lines. Be sure the replacement gets applied to all four instances.  


```{r}
#' @title Add indices for human population to parameter list
#' @description Implements [setup_MY_ix] for the newMYname model.
#' @inheritParams ramp.xds::setup_MY_ix
#' @return none
#' @importFrom utils tail
#' @export
setup_MY_ix.newMYname <- function(xds_obj, i) {with(xds_obj,{
  
  M_ix <- seq(from = max_ix+1, length.out=xds_obj$nPatches[i])
  max_ix <- tail(M_ix, 1)

  Y_ix <- seq(from = max_ix+1, length.out=xds_obj$nPatches[i])
  max_ix <- tail(Y_ix, 1)

  ... 
  
  xds_obj$max_ix = max_ix
  xds_obj$MY_obj[[i]]$ix = list(M_ix=M_ix, Y_ix=Y_ix, ...)
  return(xds_obj)
})}
```

### get_MY_vars   

In the following, each variable in your model gets its own pair of lines. 

+ Leave `H` as-is 

+ Replace `M` in `M` and `M_ix` with the first variable name. 

+ Replace `Y` in `Y` and `Y_ix` with the second variable name. 

+ Copy the form for every other variable 

+ The variable `X0` is the class that gets newborns. If you use it or if you want to have it available, compute it and add it to the returned list. 

```{r}
#' @title Return the variables as a list
#' @description This method dispatches on the type of `xds_obj$MY_obj`
#' @inheritParams ramp.xds::get_MY_vars
#' @return a [list]
#' @export
get_MY_vars.newMYname <- function(y, xds_obj, i) {
  with(xds_obj$MY_obj[[i]]$ix,{
      H = y[H_ix]
      M = y[M_ix]
      Y = y[Y_ix]
      ...
      X0 = H-M-Y 
      return(list(H=H, M=M, Y=Y,...,X0=X0))})
}
``` 

### parse_MY_orbits  

In the following, each variable in your model gets its own pair of lines. 

+ Leave `H` as-is 

+ Replace `M` in `M` and `M_ix` with the first variable name. 

+ Replace `Y` in `Y` and `Y_ix` with the second variable name. 

+ Copy the form for every other variable 

+ The variable `X0` is the class that gets newborns. If you use it or if you want to have it available, compute it and add it to the returned list.

```{r}
#' @title parse the output of deSolve and return variables for the newMYname model
#' @description Implements [parse_MY_orbits] for the newMYname model
#' @inheritParams ramp.xds::parse_MY_orbits
#' @return a named list  
#' @export
parse_MY_orbits.newMYname <- function(outputs, xds_obj, i) {
  with(xds_obj$MY_obj[[i]]$ix,{
    H <- outputs[,H_ix]
    M <- outputs[,M_ix]
    Y <- outputs[,Y_ix]
    X0 <- H-M-Y
    vars <- list(H=H, M=M, Y=Y, X0=X0)
    return(vars)
})}
```
 

## Initial Values 

### setup_MY_inits

If you did the global search and replace, you won't need to change this function. 

```{r}
#' @title Setup initial values for *newMYname*
#' 
#' @inheritParams ramp.xds::setup_MY_inits
#' 
#' @return a **`ramp.xds`** object 
#' 
#' @export
setup_MY_inits.newMYname = function(xds_obj, H, i, options=list()){
  xds_obj$MY_obj[[i]]$inits = make_MY_inits_newMYname(xds_obj$nPatches[i], H, options)
  return(xds_obj)
}
```

### make_MY_inits

```{r}
#' @title Make initial values for the Xname human model, with defaults
#' @param nPatches the number of strata in the model
#' @param options a [list] to overwrite defaults
#' @param M0 the initial value for M 
#' @param Y0 the initial value for M 
#' @return a [list]
#' @export
make_MY_inits_newMYname = function(nPatches, H, options = list(), M0=NULL, Y0=1){with(options,{
  stopifnot(is.numeric(M0))
  stopifnot(is.numeric(Y0))
  M = checkIt(M0, nPatches)
  Y = checkIt(Y0, nPatches)
  return(list(M=M, Y=Y, ...))
})}
```

### change_MY_inits 

```{r}
#' @title Return the parameters as a list
#' 
#' @inheritParams ramp.xds::change_MY_inits 
#' 
#' @return an **`xds`** object
#' @export
change_MY_inits.SIS <- function(xds_obj, i=1, options=list()) {
  with(get_MY_inits(xds_obj, i), with(options,{
    xds_obj = change_H(H, xds_obj, i)
    xds_obj$Xinits[[i]]$M = M 
    xds_obj$Xinits[[i]]$Y = Y 
    ... 
    return(xds_obj)
  }))}
```


## Dynamical Terms   

### F_fqZ  

In the following, replace ... with a formula to compute the effective density of infectious humans: it is population density by strata weighted by the probability of transmitting. 

```{r}
#' @title Size of effective infectious human population
#' @description Implements [F_fqZ] for the SIS model.
#' @inheritParams ramp.xds::F_fqZ
#' @return a [numeric] vector of length `nPatches`
#' @export
F_fqZ.newMYname <- function(y, xds_obj, i) {
  with(get_MY_vars(y, xds_obj, i), 
    with(xds_obj$MY_obj[[i]], {
      fqZ = ...  
      return(fqZ)
}))}
```

### F_fqM  

In the following, replace ... with a formula to compute the effective density of infectious humans: it is population density by strata weighted by the probability of transmitting. 

```{r}
#' @title Size of effective infectious human population
#' @description Implements [F_fqM] for the SIS model.
#' @inheritParams ramp.xds::F_fqM
#' @return a [numeric] vector of length `nPatches`
#' @export
F_fqM.newMYname <- function(y, xds_obj, i) {
  with(get_MY_vars(y, xds_obj, i), 
    with(xds_obj$MY_obj[[i]], {
      fqM = ...  
      return(fqM)
}))}
```

### F_eggs


```{r}
#' @title Infection blocking pre-erythrocytic immunity
#' @description Implements [F_eggs] for the newMYname model.
#' @inheritParams ramp.xds::F_eggs
#' @return a [numeric] vector of length `nPatches`
#' @export
F_eggs.newMYname <- function(y, xds_obj, i) {
  with(xds_obj$MY_obj[[i]],{ 
    ########################
    # retrieve or compute it 
    ########################
    eggs = ... 
    ########################
    # return it 
    ########################
    return(eggs)
  })
}
```


## Other Outputs  

To make `ramp.xds` easy to use, we provide some functions that parse the outputs and compute some standard metrics. 

### F_y?  

## Internal Consistency

## Optional Functions

### Plotting   

### 16. xds_lines_X   

#### xds_lines_X_SIS

```{r, eval=F, purl=F}
#' Add lines for the density of infected individuals for the SIS model
#'
#' @param MY a list with the outputs of parse_deout_X_SIS
#' @param nPatches the number of population strata
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#'
#' @export
xds_lines_X_SIS = function(MY, nPatches, clrs=c("darkblue","darkred"), llty=1){
  with(MY,{
    if(nPatches==1) {
      lines(time, S, col=clrs[1], lty = llty[1])
      lines(time, I, col=clrs[2], lty = llty[1])
    }
    if(nPatches>1){
      if (length(clrs)==2) clrs=matrix(clrs, 2, nPatches)
      if (length(llty)==1) llty=rep(llty, nPatches)

      for(i in 1:nPatches){
        lines(time, S[,i], col=clrs[1,i], lty = llty[i])
        lines(time, I[,i], col=clrs[2,i], lty = llty[i])
      }
    }
  })}
```

#### xds_lines_X_newMYname

```{r}
#' Add lines for the density of infected individuals for the newMYname model
#'
#' @param MY a list with the outputs of parse_deout_X_newMYname
#' @param nPatches the number of population strata
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#'
#' @export
xds_lines_X_newMYname = function(MY, nPatches, clrs=c("darkblue","darkred"), llty=1){
  with(MY,{
    if(nPatches==1) {
      lines(time, S, col=clrs[1], lty = llty[1])
      lines(time, I, col=clrs[2], lty = llty[1])
    }
    if(nPatches>1){
      if (length(clrs)==2) clrs=matrix(clrs, 2, nPatches)
      if (length(llty)==1) llty=rep(llty, nPatches)

      for(i in 1:nPatches){
        lines(time, S[,i], col=clrs[1,i], lty = llty[i])
        lines(time, I[,i], col=clrs[2,i], lty = llty[i])
      }
    }
  })}
```

### 17. xds_plot_X   

We provide a basic plotting function. The function `xds_lines_X_*` does not have an `S3` version. Instead, it is called by the `S3` function `xds_plot_X`

#### Generic

```{r, eval=F, purl=F}
#' Basic plotting for epidemiological models
#'
#' @param xds_obj a list that defines an `ramp.xds` model (*e.g.*,  generated by `xde_setup()`)
#' @param i the host species index
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#' @param stable a logical: set to FALSE for `orbits` and TRUE for `stable_orbits`
#' @param add_axes a logical: plot axes only if TRUE
#'
#' @export
xds_plot_X = function(xds_obj, i=1, clrs="black", llty=1, stable=FALSE, add_axes=TRUE){
  UseMethod("xds_plot_X", xds_obj$MY_obj[[i]])
}
```

#### xds_plot_X.SIS

```{r, eval=F, purl=F}
#' Plot the density of infected individuals for the newMYname model
#'
#' @inheritParams ramp.xds::xds_plot_X
#' @export
xds_plot_X.SIS = function(xds_obj, i=1, clrs=c("darkblue","darkred"), llty=1, stable=FALSE, add_axes=TRUE){
  vars=with(xds_obj$outputs,if(stable==TRUE){stable_orbits}else{orbits})

  if(add_axes==TRUE)
    with(vars$MY[[i]],
         plot(time, 0*time, type = "n", ylim = c(0, max(H)),
              ylab = "# Infected", xlab = "Time"))


  xds_lines_X_SIS(vars$MY[[i]], xds_obj$Hpar[[i]]$nPatches, clrs, llty)
}
```

#### xds_plot_X.newMYname

```{r}
#' Plot the density of infected individuals for the newMYname model
#'
#' @inheritParams ramp.xds::xds_plot_X
#' @export
xds_plot_X.newMYname = function(xds_obj, i=1, clrs=c("darkblue","darkred"), llty=1, stable=FALSE, add_axes=TRUE){
  vars=with(xds_obj$outputs,if(stable==TRUE){stable_orbits}else{orbits})

  if(add_axes==TRUE)
    with(vars$MY[[i]],
         plot(time, 0*time, type = "n", ylim = c(0, max(H)),
              ylab = "# Infected", xlab = "Time"))

  xds_lines_X_newMYname(vars$MY[[i]], xds_obj$Hpar[[i]]$nPatches, clrs, llty)
}
```


## Reusable Code 

**`ramp.xds`** has a modular design. The modular design makes it possible to break down published models to serve as dynamical components in new models for malaria, a kind of plug-and-play functionality for two kinds of systems: 

+ systems of ordinary and delay differential equations models for malaria and other mosquito-transmitted pathogens. 

+ stochastic and deterministic discrete time systems for malaria and other mosquito-transmitted pathogens. 

This package -- `ramp.library` -- is a collection of models to implement a large number of dynamical model families and other algorithms taken from the literature describing malaria and other mosquito-transmitted pathogens (see Reiner, *et al.* 2013)^[Reiner RC Jr, Perkins TA, Barker CM, Niu T, Chaves LF, Ellis AM, et al. A systematic review of mathematical models of mosquito-borne pathogen transmission: 1970-2010. J R Soc Interface. 2013;10: 20120921.]. Those models have been implemented here as code that has undergone rigorous testing and that can be used to build models for simulation-based analytics.  

The mathematics supporting this framework were explained in [Spatial Dynamics of Malaria Transmission](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010684){target="_blank"}
^[Wu SL, Henry JM, Citron DT, Mbabazi Ssebuliba D, Nakakawa Nsumba J, Sánchez C. HM, et al. (2023) Spatial dynamics of malaria transmission. PLoS Comput Biol 19(6): e1010684. https://doi.org/10.1371/journal.pcbi.1010684] 

This vignette was written to help add new modules to `ramp.library.` 

