---
title: "The Garki Model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Garki Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, echo=FALSE, warning=FALSE, messages=FALSE, purl=F}
suppressMessages(library(knitr))
suppressMessages(library(ramp.xds))
suppressMessages(library(ramp.library))
```


```{r, fig.width=6, fig.height=3.5}
garki <- xds_setup(Xname = "garki", MYZname = "si")
garki <- xde_solve(garki, Tmax=3650, dt = 15)
xds_plot_X(garki, 1)
```


In the early 1950s, George Macdonald's updated Ross's mathematical theory and published a quantitative synthesis of malaria epidemiology and control. In field trials in the late 1960s, Macdonald's models were tested, exposing some of their weaknesses (Nájera, 1973; Nájera, 1974). Many of these weaknesses were addressed when Klaus Dietz and his colleagues published a new malaria model, in the 1970s, that was field tested in The Garki Project (Dietz, Molineaux, & Thomas, 1974). 

In ramp.xds, we have coded the Garki model in two ways: 1. a discrete-time system, as it was originally written; and 2. as a continuous time system. 

***

References: 

1. Nájera , J. A & World Health Organization. (1973). A critical review of the field application of a mathematical model of malaria epidemiology. Available from the [World Health Organization](https://iris.who.int/bitstream/handle/10665/65662/WHO_MAL_73.796.pdf)

2.  Nájera JA. A critical review of the field application of a mathematical model of malaria eradication. Bull World Health Organ. 1974;50: 449–457. Available from [Bull. WHO](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2481134/pdf/bullwho00166-0077.pdf)

3. Dietz K, Molineaux L, Thomas A. A malaria model tested in the African savannah. Bull World Health Organ. 1974;50: 347–357. Available from [Bull. WHO](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2481204/pdf/bullwho00165-0208.pdf). 
