---
title: "Adding new XH models to `ramp.xde`" 
subtitle: "Models for Human / Vertebrate Host Infection Dynamics" 
output:
  html_document: 
    theme: paper 
vignette: >
  %\VignetteIndexEntry{Adding new XH models to `ramp.xde`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*** 

This vignette discusses how to add a new model family to [`ramp.library`](https://github.com/dd-harp/ramp.library) that can be used by  [`ramp.xde`](https://github.com/dd-harp/ramp.xde). This vignette covers models of parasite infection dynamics and immunity in humans or other hosts -- the dynamical components we have denoted $\cal X$  and $\cal H.$

*** 

We have written this vignette to produce a new template file using `knitr::purl,` which can be modified to build a new model. Simply execute the following command, to create `Xmod-SIRnew.R.`  Then, rename `Xmod-SIRnew.R` by replacing `SIRnew` in the filename. Then open the file and do a global search for `SIRnew` and replace it with your new name. 

```{r, purl=F, eval=T}
knitr::purl("X-How_To_Add_Model.Rmd", 
            output = "Xmod-SIRnew.R", 
            documentation = 0, 
            quiet=TRUE)
```

This is a good start, but there's more work to do.
For more details, read the text that follows.

This basic introduction is lacking some information that typical users might find useful. In particular, we have defined best practices encoding the $\cal H$ component of models. A vignette is planned.  


### Reusable Code 

This package -- `ramp.library` -- contains reusable code that has been rigorously tested and that implements a large number of dynamical model families and other algorithms taken from the literature describing malaria and other mosquito-transmitted pathogens (see Reiner, *et al.* 2013)^[Reiner RC Jr, Perkins TA, Barker CM, Niu T, Chaves LF, Ellis AM, et al. A systematic review of mathematical models of mosquito-borne pathogen transmission: 1970-2010. J R Soc Interface. 2013;10: 20120921.]. The supporting code was designed to be modular, and plug-and-play. The modular design makes it possible to break down published models to serve as the dynamical components in new models for malaria. The model families in `ramp.library` supports nimble model building that can be used by:

+ [`ramp.xde`](https://github.com/dd-harp/ramp.xde) implements a modular, flexible, and extensible framework for building systems of ordinary and delay differential equations models for malaria and other mosquito-transmitted pathogens. The mathematic framework for `ramp.xde` was explained in [Spatial Dynamics of Malaria Transmission](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010684){target="_blank"}
^[Wu SL, Henry JM, Citron DT, Mbabazi Ssebuliba D, Nakakawa Nsumba J, Sánchez C. HM, et al. (2023) Spatial dynamics of malaria transmission. PLoS Comput Biol 19(6): e1010684. https://doi.org/10.1371/journal.pcbi.1010684] 

+ [`ramp.dts`](https://github.com/dd-harp/ramp.dts) implements a modular, flexible, and extensible framework for building discrete time systems, including both deterministic and stochastic difference equations for malaria and other mosquito-transmitted pathogens. 

This vignette was written to help add new models to `ramp.xde.` 

In planned companion to this vignette, we show how to add models of $\cal XH$ to [`ramp.dts`](https://github.com/dd-harp/ramp.dts). 

***


### Functions 

A new model family of class $\cal XH$ -- human infection dynamics, including immunity -- 
has 17 required functions encoded using a standard template. Some of these required functions are `S3` class functions in [human-interface.R](https://github.com/dd-harp/ramp.xde/blob/main/R/human-interface.R){target="_blank"}, and several examples are posted in the `ramp.xde` github repository, including the `SIS` model [human-SIS.R](https://github.com/dd-harp/ramp.xde/blob/main/R/human-SIS.R){target="_blank"}. 

To configure a new model, you must choose a string called `SIRnew`. The only rule for choosing `SIRnew` is that it must be unique. No other ${\cal XH}$ class model in `ramp.xde` or `ramp.library` can be using that string already. All the `S3` functions get dispatched by the string that replaces `SIRnew.` 

The 17 required functions are:

**Derivatives** 

1. `dXdt.SIRnew` - differential equations are defined by a system of differential equations. in `ramp.xde` these are encoded in a function called `dXdt` that  computes and returns the **derivatives.**  The function is set up so that the system can be solved by `deSolve::ode` or `deSolve::dede.` 

**Variables** - A set of functions is set up that document the variables by name and assign each one an index, to store, assign and update initial values. These functions streamline setup, and guarantee consistency. 

2. `make_Xinits_SIRnew` - each model requires a function that documents the variables by name and assigns initial values 
    
3. `setup_Xinits.SIRnew` - is a wrapper, that gets called by `ramp.xde::xde_setup.` It gets called by `make_Xinits_SIRnew,` and for the $i^{th}$ host, the initial values are stored as `pars$Xinits[[i]].` 

4. `make_indices_X.SIRnew` - is the function that assigns an index to each variable in the model, and stores it as `pars$ix$X[[i]].` The indices are returned as a named list.  

5. `list_Xvars.SIRnew` - retrieves the value of variables at a point in time and returns the values by name in a list; the function gets called by `dXdt` and by `update_inits_X` and it can be useful in other contexts. 

6. `get_inits_X.SIRnew` - retrieves the stored initial values in the same order that they are returned by `dXdt.SIRnew`

7. `update_inits_X.SIRnew` - a utility that makes it possible to extract and store a new set of initial values for `SIRnew` from a vector $y$ holding all the variables in a model. 

**Parameters**

8. `make_Xpar_SIRnew` returns a formatted list: 

    + the parameter values are stored by name 
   
    + `class(pars)` = `SIRnew` 
    
9. `setup_Xpar.SIRnew` is a wrapper that calls `make_Xpar_SIRnew` 

**Dynamical Terms** are Computed using a Standard Set of Functions 

10. `F_X.SIRnew` - compute the effective infective density of the vertebrate hosts 

11. `F_H.SIRnew` - compute the human population density for the host strata

12. `F_b.SIRnew` - compute the probability a most will become infected after receiving an infective bite, which combines elements of this $\cal X$ model with a model of environmental heterogeneity in `Exposure.` 

**Outputs**

13. `parse_deout_X.SIRnew` - this function parses the output of `deSolve` and returns a named list that is either a vector or a matrix holding the values of the variables at each point in time. It works like `list_Xvars,` which accepts a vector holding $y$, the values of the variables at a single point in time. 

14. `F_pr.SIRnew` - this function takes 

15. `HTC.SIRnew`

**Outputs**

16. `xde_lines_X.SIRnew` 

17. `xde_plot_X.SIRnew` is a wrapper that calls `xde_lines_X` 

### Style 

We recommend that new models conform to a set of style guidelines:

+ All the names of the variables by name using the indices attached to `Xpar` 

+ All documentation is written in `roxygen2`

+ Functions should use `with(*,{...})` to make the code easier to read. 

+ A vignette should accompany each model, including citations to the relevant publications, and proposed tests of the code.  

+ The functions should easy to relate to the equations defining a model, following notation defined in an accompanying vignette.


## Derivatives  

### dXdt 

The function `dXdt` is established as a generic function that dispatches on `Xpar.` Since there might be many host species, it actually dispatches on the $i^{th}$ element in a list, `pars$Xpar[[i]].`

```{r, purl=F, eval=F}
#' @title Derivatives for human population
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param t current simulation time
#' @param y state vector
#' @param pars a list
#' @param i the host species index
#' @return a [numeric] vector
#' @export
dXdt <- function(t, y, pars, i) {
  UseMethod("dXdt", pars$Xpar[[i]])
}
```

### dXdt.SIS

This is a version of the `SIS` model in `ramp.xde` with demographics. If a model has any demographic processes (*e.g.* aging, migration, births and deaths), then each variable's derivative must be transformed by the same matrix as the human population density, `H`. Any disease induced mortality should be included in a matrix that computes changes in human population density, `dHdt`. 

In the following, we use the `with()` function to expose parameter names attached to lists to make the code easy to read. 

```{r, purl=F, eval=F}
#' @title Derivatives for human population
#' @description Implements [dXdt] for the SIS model, no demography.
#' @inheritParams dXdt
#' @return a [numeric] vector
#' @export
dXdt.SIS <- function(t, y, pars, i) {

  foi <- pars$FoI[[i]]

  with(list_Xvars(y, pars, i),{
    H <- F_H(t, y, pars, i)
    with(pars$Xpar[[i]], {
      dS <- Births(t, H, pars, i) - foi*S + r*I + dHdt(t, S, pars, i)
      dI <- foi*S - r*I + dHdt(t, I, pars, i)
      return(c(dS, dI))
    })
  })
}
```


### dXdt.SIRnew 

This *should* be straightforward, but there's one very important wrinkle. If you want to be able to use the `ramp.xde` capabilities for cohort dynamics and vital dynamics, then the dynamics must be extended carefully to call the demographic matrix function `dHdt.` This can become tricky, so we handle it elsewhere. When writing the code:

+ The foi gets computed in `Exposure` and stored as `pars$FoI[[i]]` for the $i^{th}$ host species. 

+ In the next step, the variables are extracted from the variable vector, `y`, using the indices, which are stored in `pars$ix$X[[i]]` 

+ We compute the derivatives. 

+ We return the variables in order. Be sure to establish and then maintain a convention for ordering the variables in your model, and then maintain that order in every function. In particular: 

    - set up the initial values in that order 
    
    - return the derivatives in that order

+ Be careful about how you nest the `with(*, ...)` calls. 


```{r}
#' @title Compute the derivatives for parasite infection dynamics in human population strata 
#' @description Implements [dXdt] for the SIRnew model
#' @inheritParams ramp.xde::dXdt
#' @return a [numeric] vector
#' @export
dXdt.SIRnew<- function(t, y, pars, i) {
  # do not change this
  foi <- pars$FoI[[i]]
  
  # attach the variables by name
  with(list_Xvars(y, pars, i),{
    # compute H (if it isn't one of the variables)
    H <- F_H(t, y, pars, i)
    
    # expose the parameters (see make_Xpar_SIRnew)
    with(pars$Xpar[[i]], {
      # compute the derivatives
      dS <- Births(t, H, pars,i) + dHdt(t, S, pars,i) - foi*S 
      dI <- foi*S- r*I + dHdt(t, I, pars,i)
      dR <- r*I + dHdt(t, R, pars,i)
      
      # concatenate the derivatives
      derivs = c(dS, dI, dR)
      
      # return the derivatives
      return(derivs)
    })
  })
}
```

## Variables  

### 2. make_Xinits  

#### make_Xinits_SIS

```{r, eval=F, purl=F}
#' @title Make initial values for the SIS human model, with defaults
#' @param nStrata the number of strata in the model
#' @param Xopts a [list] to overwrite defaults
#' @param H0 the initial human population density
#' @param S0 the initial values of the parameter S
#' @param I0 the initial values of the parameter I
#' @return a [list]
#' @export
make_Xinits_SIS = function(nStrata, Xopts = list(), H0=NULL, S0=NULL, I0=1){with(Xopts,{
  if(is.null(S0)) S0 = H0 - I0
  stopifnot(is.numeric(S0))
  S = checkIt(S0, nStrata)
  I = checkIt(I0, nStrata)
  return(list(S=S, I=I))
})}
```

#### make_Xinits_SIRnew

```{r}
#' @title Make initial values for the SIRnew human model, with defaults
#' @param nStrata the number of strata in the model
#' @param Xopts a [list] to overwrite defaults
#' @param H0 the initial value for H
#' @param S0 the initial value for S
#' @param I0 the initial value for I
#' @param R0 the initial values for R
#' @return a [list]
#' @export
make_Xinits_SIRnew = function(nStrata, Xopts = list(), H0= NULL, S0=NULL, I0=1, R0 = 1){with(Xopts,{
  if(is.null(S0)) S0 = H0-(I0+R0)
  stopifnot(is.numeric(S0))
  S = checkIt(S0, nStrata)
  I = checkIt(I0, nStrata)
  R = checkIt(R0, nStrata)
  return(list(S=S, I=I, R =R))
})}

```

### 3. setup_Xinits  

In the `xde_setup` family of setup functions, initial values for a model fulfilling the dynamical component $\cal X,$ called `SIRnew` is set up by the function `setup_Xinits.` 

#### Generic 

The `setup_Xinits` function dispatches on `pars$Xpar[[i]]`

```{r, eval=F, purl=F}
#' @title A function to set up Xpar
#' @description This method dispatches on `SIRnew`.
#' @param pars a [list]
#' @param i the host species index
#' @param Xopts a [list]
#' @return a [list]
#' @export
setup_Xinits = function(pars, i, Xopts=list()){
  UseMethod("setup_Xinits", pars$Xpar[[i]])
}
```

#### setup_Xinits.SIS

The `setup_Xinits.SIS` calls `make_Xinits_SIS` 

```{r, eval=F, purl=F}
#' @title Setup Xinits.SIS
#' @description Implements [setup_Xinits] for the SIS model
#' @inheritParams setup_Xinits
#' @return a [list] vector
#' @export
setup_Xinits.SIS = function(pars, i, Xopts=list()){
  pars$Xinits[[i]] = with(pars, make_Xinits_SIS(pars$Hpar[[i]]$nStrata, Xopts, H0=Hpar[[i]]$H))
  return(pars)
}
```

#### setup_Xinits.SIRnew

To modify, replace `SIRnew` 

```{r}
#' @title Setup Xinits.SIRnew
#' @description Implements [setup_Xinits] for the SIRnew model
#' @inheritParams ramp.xde::setup_Xinits
#' @return a [list] vector
#' @export
setup_Xinits.SIRnew = function(pars, i, Xopts=list()){
  pars$Xinits[[i]] = with(pars, make_Xinits_SIRnew(pars$Hpar[[i]]$nStrata, Xopts, H0=Hpar[[i]]$H))
  return(pars)
}
```

### 4. make_indices_X  

#### Generic

```{r, eval=F, purl=F}
#' @title Add indices for human population to parameter list
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param pars a [list]
#' @param i the host species index
#' @return a [list]
#' @export
make_indices_X <- function(pars, i) {
  UseMethod("make_indices_X", pars$Xpar[[i]])
}
``` 

#### make_indices_X.SIS

```{r, eval=F, purl=F}
#' @title Add indices for human population to parameter list
#' @description Implements [make_indices_X] for the SIS model.
#' @inheritParams ramp.xde::make_indices_X
#' @return none
#' @importFrom utils tail
#' @export
make_indices_X.SIS <- function(pars, i) {with(pars,{

  S_ix <- seq(from = max_ix+1, length.out=Hpar[[i]]$nStrata)
  max_ix <- tail(S_ix, 1)

  I_ix <- seq(from = max_ix+1, length.out=Hpar[[i]]$nStrata)
  max_ix <- tail(I_ix, 1)

  pars$max_ix = max_ix
  pars$ix$X[[i]] = list(S_ix=S_ix, I_ix=I_ix)
  return(pars)
})}
```

#### make_indices_X.SIRnew

```{r}
#' @title Add indices for human population to parameter list
#' @description Implements [make_indices_X] for the SIRnew model.
#' @inheritParams ramp.xde::make_indices_X
#' @return none
#' @importFrom utils tail
#' @export
make_indices_X.SIRnew <- function(pars, i) {with(pars,{
  
  S_ix <- seq(from = max_ix+1, length.out=Hpar[[i]]$nStrata)
  max_ix <- tail(S_ix, 1)
  
  I_ix <- seq(from = max_ix+1, length.out=Hpar[[i]]$nStrata)
  max_ix <- tail(I_ix, 1)
  
  R_ix <- seq(from = max_ix+1, length.out=Hpar[[i]]$nStrata)
  max_ix <- tail(R_ix, 1)
  
  
  
  pars$max_ix = max_ix
  pars$ix$X[[i]] = list(S_ix=S_ix, I_ix=I_ix, R_ix=R_ix)
  return(pars)
})}
```


### 5. list_Xvars   

#### Generic

```{r, eval=F, purl=F}
#' @title Return the variables as a list
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param y the variables
#' @param pars a [list]
#' @param i the host species index
#' @return a [list]
#' @export
list_Xvars <- function(y, pars, i) {
  UseMethod("list_Xvars", pars$Xpar[[i]])
}
``` 

#### list_Xvars.SIS

```{r, eval=F, purl=F}
#' @title Return the variables as a list
#' @description This method dispatches on the type of `pars$Xpar`
#' @inheritParams list_Xvars
#' @return a [list]
#' @export
list_Xvars.SIS <- function(y, pars, i) {
  with(pars$ix$X[[i]],
      return(list(
        S = y[S_ix],
        I = y[I_ix]
      )
  ))
}
``` 

#### list_Xvars.SIRnew 

```{r}
#' @title Return the variables as a list
#' @description This method dispatches on the type of `pars$Xpar`
#' @inheritParams ramp.xde::list_Xvars
#' @return a [list]
#' @export
list_Xvars.SIRnew <- function(y, pars, i) {
  with(pars$ix$X[[i]],
       return(list(
         S = y[S_ix],
         I = y[I_ix],
         R = y[R_ix]
       )
       ))
}
``` 


### 6. get_inits_X   

By now the pattern should be obvious. 

#### Generic 

```{r, eval=F, purl=F}
#' @title Return initial values as a vector
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param pars a [list]
#' @param i the host species index
#' @return none
#' @export
get_inits_X <- function(pars, i) {
  UseMethod("get_inits_X", pars$Xpar[[i]])
}
```

#### get_inits_X.SIS

Returns the initial values, in order, by name. 

```{r, eval=F, purl=F}
#' @title Return initial values as a vector
#' @description This method dispatches on the type of `pars$Xpar`.
#' @inheritParams get_inits_X
#' @return a [numeric] vector
#' @export
get_inits_X.SIS <- function(pars, i){
  with(pars$Xinits[[i]], return(c(S,I)))
}
```

#### get_inits_X.SIRnew

Write the function to return the initial values, in order, by name. 

```{r}
#' @title Return initial values as a vector
#' @description This method dispatches on the type of `pars$Xpar`.
#' @inheritParams ramp.xde::get_inits_X
#' @return a [numeric] vector
#' @export
get_inits_X.SIRnew <- function(pars, i){
  with(pars$Xinits[[i]], return(c(S,I,R)))
}

```

### 7. update_inits_X  

Pass a vector `y`, extract the values of the variables, and modify the initial values.

#### Generic 

```{r, purl=F, eval=F}
#' @title Set the initial values from a vector of states
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param pars a [list]
#' @param y0 a vector of initial values
#' @param i the host species index
#' @return none
#' @export
update_inits_X <- function(pars, y0, i) {
  UseMethod("update_inits_X", pars$Xpar[[i]])
}
```

#### update_inits_X.SIS

```{r, purl=F, eval=F}
#' @title Update inits for the SIS human model from a vector of states
#' @inheritParams update_inits_X
#' @return none
#' @export
update_inits_X.SIS <- function(pars, y0, i) {
  with(list_Xvars(y0, pars, i),{
    pars$Xinits[[i]] = make_Xinits_SIS(pars, list(), S0=S, I0=I)
    return(pars)
})}
```

#### update_inits_X.SIRnew

```{r}
#' @title Update inits for the SIRnew human model from a vector of states
#' @inheritParams ramp.xde::update_inits_X
#' @return none
#' @export
update_inits_X.SIRnew <- function(pars, y0, i) {
  with(list_Xvars(y0, pars, i),{
    pars = make_Xinits_SIRnew(pars, list(), S0=S, I0=I, R0=R)
    return(pars)
  })}
```


## Parameters  

The second function sets up a list `Xpar` that dispatches `dXdt` and that has the parameter values. 

### 8. make_Xpar_SIS  

There is no `S3` definition for the `make_Xpar_*` class of functions. Instead, we write a generic `S3` function called `setup_Xpar` that calls a function `make_Xpar_*` There is, nevertheless, a reasonably standard formula for `make_Xpar_*`  

+ The first argument is `nStrata` the number of population strata for the species. 

+ The first argument is `Xopts().`  Notice that `Xopts()` is an empty list by default, but if a non-empty list is passed, the values overwrite the defaults. 

+ The parameters are attached by name. These parameter names must be the same ones used in `dXdt`

+ By default, they should have `length(*)= nStrata`

+ The last part of the function name is the string that dispatches the `S3` functions. In `make_Xpar_SIS,` we write `class(Xpar) <- "SIS"` 

#### make_Xpar_SIS

```{r, eval=F, purl=F}
#' @title Make parameters for SIS human model, with defaults
#' @param nStrata is the number of population strata
#' @param Xopts a [list] that could overwrite defaults
#' @param b transmission probability (efficiency) from mosquito to human
#' @param c transmission probability (efficiency) from human to mosquito
#' @param r recovery rate
#' @return a [list]
#' @export
make_Xpar_SIS = function(nStrata, Xopts=list(),
                         b=0.55, r=1/180, c=0.15){
  with(Xopts,{
    Xpar = list()
    class(Xpar) <- "SIS"

    Xpar$b = checkIt(b, nStrata)
    Xpar$c = checkIt(c, nStrata)
    Xpar$r = checkIt(r, nStrata)

    return(Xpar)
  })}
```

#### make_Xpar_SIRnew

To format this for a new function, do a search and replace on `SIRnew` and replace the parmeter names and values $p1$, $p2$, $\ldots$ 

```{r}
#' @title Make parameters for SIRnew human model, with defaults
#' @param nStrata is the number of population strata
#' @param Xopts a [list] that could overwrite defaults
#' @param b the proportion of infective bites that cause an infection
#' @param r the the duration of an infection
#' @param c the proportion of bites on infected humans that infect a mosquito
#' @return a [list]
#' @export
make_Xpar_SIRnew = function(nStrata, Xopts=list(),
                          b=0.55, r=1/180, c=0.15){
  with(Xopts,{
    Xpar = list()
    class(Xpar) <- c("SIRnew")
    
    Xpar$b = checkIt(b, nStrata)
    Xpar$c = checkIt(c, nStrata)
    Xpar$r = checkIt(r, nStrata)
    
    return(Xpar)
  })}
```


### 9. setup_Xpar  

In the `xde_setup` family of setup functions, setup for a model fulfilling the dynamical component $\cal X,$ called `SIRnew` is set up by the function `setup_Xpar.` 

#### Generic 

This function must set up `Xpar` and it must also set up the initial values. By convention, the functions that get called are not `S3` class functions.   

The `setup_Xpar` functions assigns to `Xname` the `class(Xname)` and dispatches on `Xname`

```{r, eval=F, purl=F}
#' @title A function to set up Xpar
#' @description This method dispatches on `Xname`.
#' @param Xname a [character] string
#' @param pars a [list]
#' @param i the host species index
#' @param Xopts a [list]
#' @return a [list]
#' @export
setup_Xpar = function(Xname, pars, i, Xopts=list()){
  class(Xname) <- Xname
  UseMethod("setup_Xpar", Xname)
}
```

#### setup_Xpar.SIS

The dispatched function `SIS` ignores xname, but it calls a function `make_Xpar_SIS` 

```{r, eval=F, purl=F}
#' @title Setup Xpar.SIS
#' @description Implements [setup_Xpar] for the SIS model
#' @inheritParams setup_Xpar
#' @return a [list] vector
#' @export
setup_Xpar.SIS = function(Xname, pars, i, Xopts=list()){
  pars$Xpar[[i]] = make_Xpar_SIS(pars$Hpar[[i]]$nStrata, Xopts)
  return(pars)
}
```

#### setup_Xpar.Xname

To use this, simply to a search and replace on `Xname.` 

```{r}
#' @title Setup Xpar.SIRnew
#' @description Implements [setup_Xpar] for the SIRnew model
#' @inheritParams ramp.xde::setup_Xpar
#' @return a [list] vector
#' @export
setup_Xpar.SIRnew = function(Xname, pars, i, Xopts=list()){
  pars$Xpar[[i]] = make_Xpar_SIRnew(pars$Hpar[[i]]$nStrata, Xopts)
  return(pars)
}
```



## Terms   

All models in `ramp.xde` the modular design dictates that we must define a few functions that compute a few key quantities: 

+ `X` is defined to be the effective density of infectious humans, retrieved by `F_X`

+ `H` is defined to be the size of each population stratum, retrieved by `F_H`

+ `b` is defined to be the fraction of infective bites that cause an infection, retrieved by `F_b`

### 10. F_X  

The function `F_X` returns the effective density of infectious humans for each stratum. The notion is that $X$ is proportional to $H$, where the constant is the fraction of bites on the strata that would infect a mosquito.  

#### Generic 

```{r, purl=F, eval=F}
#' @title Size of effective infectious human population
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param t current simulation time
#' @param y state vector
#' @param pars a list
#' @param i the host species index
#' @return a [numeric] vector of length `nStrata`
#' @export
F_X <- function(t, y, pars, i) {
  UseMethod("F_X", pars$Xpar[[i]])
}
```

#### F_X.SIS 

As an example, we include the $SIS$ model from `ramp.xde.` In the `SIS` model, the returned value is $cI$ where $I$ is the density of infected individuals; the prevalence of infection is $X/H$. The constant $c$ is the probability a mosquito would become infected after blood feeding on an infected human. Two things to note: 

+ First, `ramp.xde` holds the indices by component under `pars$ix,` so `pars$ix$X` holds the indices for all the variables for all species, and `pars$ix$X[[i]]` holds the indices for the $i^{th}$ species. Putting them under `X` helps to avoid conflicts if, for example, another dynamical component happened to use the same variable name.  



```{r, purl=F, eval=F}
#' @title Size of effective infectious human population
#' @description Implements [F_X] for the SIS model.
#' @inheritParams F_X
#' @return a [numeric] vector of length `nStrata`
#' @export
F_X.SIS <- function(t, y, pars, i) {
  I = y[pars$ix$X[[i]]$I_ix]
  X = with(pars$Xpar[[i]], c*I)
  return(X)
}
```

The commented code relevant code that must be changed is commented out. The preferred style is to *extract* the relevant variables by name, then *compute* the effective infectious density,or individuals in this stratum, then return the value. In the manuscript$^1$, this quantity was called **X.** 

#### F_X.SIRnew

The following is a template for new model. The declaration of `inheritParams` from the `SIS` model in `ramp.xde` works naturally. The namespace from `ramp.xde` is needed, so our SIRnew function (above) has `ramp.xde::F_X.` 

The steps are to extract the variables in 

```{r}
#' @title Size of effective infectious human population
#' @description Implements [F_X] for the SIS model.
#' @inheritParams ramp.xde::F_X
#' @return a [numeric] vector of length `nStrata`
#' @export
F_X.SIRnew <- function(t, y, pars, i) {
  I = y[pars$ix$X[[i]]$I_ix]
  Y = with(pars$Xpar[[i]], c*I)
  return(Y)
}
```

### 11. F_H  

The total population size, $H$, is often important, so we write a function that computes it.   

#### Generic 

```{r, purl=F, eval=F}
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param t current simulation time
#' @param y state vector
#' @param pars a list
#' @param i the host species index
#' @return a [numeric] vector of length `nStrata`
#' @export
F_H <- function(t, y, pars, i) {
  UseMethod("F_H", pars$Xpar[[i]])
}
```

#### F_H.SIS

This is the function that computes $H = S+I$ for the `SIS` model. 

```{r, purl=F, eval=F}
#' @title Size of effective infectious human population
#' @description Implements [F_H] for the SIS model.
#' @inheritParams F_H
#' @return a [numeric] vector of length `nStrata`
#' @export
F_H.SIS <- function(t, y, pars, i){
  with(list_Xvars(y, pars, i), return(S+I))
}
```

#### F_H.SIRnew

The function `F_X` returns the effective density of infectious humans for each stratum. 
The following is a template for a model: 

```{r}
#' @title Size of effective infectious human population
#' @description Implements [F_H] for the SIRnew model.
#' @inheritParams ramp.xde::F_H
#' @return a [numeric] vector of length `nStrata`
#' @export
F_H.SIRnew <- function(t, y, pars, i){
  with(list_Xvars(y, pars, i), {
    H <- S + I+R
    return(H)
  })}
```

### 12. F_b  

We use the short parameter name `b` to denote the fraction of infective bites by mosquitoes that cause an infection. This is computed in another function, called `Exposure` under a model for environmental heterogeneity that computes a local FoI based on the local daily EIR. Then, the total FoI is computed from a travel model.  

#### Generic 

```{r, purl=F, eval=F}
#' @title Infection blocking pre-erythrocytic immunity
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param y state vector
#' @param pars a list
#' @param i the host species index
#' @return a [numeric] vector of length `nStrata`
#' @export
F_b <- function(y, pars, i) {
  UseMethod("F_b", pars$Xpar[[i]])
}
```

#### `F_b.SIS`

In `ramp.xde,` we recognize the problem of translating the local EIR into the FoI. The local EIR is computed as one of the core model terms for each one of the model strata. 

The function `F_b` returns the model-defined probability an infective bite by a mosquito would cause an infection. This could include the combined effects of inefficient transmission, pre-erytrocytic immunity, and perhaps some effects of blood stage immunity. In a model with no effect, the return value would be $1$.  

Once again, we provide the `SIS` model from `ramp.xde` as our example: 

```{r, purl=F, eval=F}
#' @title Infection blocking pre-erythrocytic immunity
#' @description Implements [F_b] for the SIS model.
#' @inheritParams F_b
#' @return a [numeric] vector of length `nStrata`
#' @export
F_b.SIS <- function(y, pars, i) {
  return(with(pars$Xpar[[i]], b))
}
```

#### F_b.SIRnew 

Depending on the model, the value of `b` is either retrieved or it is computed from other variables in the model, and it is expected that `length(b) = nStrata` 

```{r}
#' @title Infection blocking pre-erythrocytic immunity
#' @description Implements [F_b] for the SIRnew model.
#' @inheritParams ramp.xde::F_b
#' @return a [numeric] vector of length `nStrata`
#' @export
F_b.SIRnew <- function(y, pars, i) {
  with(pars$Xpar[[i]],{ 
    ########################
    # retrieve or compute it 
    ########################
    b = pars$Xpar[[i]]$b
    ########################
    # return it 
    ########################
    return(b)
  })
}
```




## Outputs  

To make `ramp.xde` easy to use, we provide some functions that parse the outputs and compute some standard metrics. 

### 13. `parse_deout_X`  

`parse_deout_X` takes the output of `deSolve,` parses the outputs using the variable indices, and returns the variables by name as a list. 

#### Generic 

```{r, eval=F, purl=F}
#' @title Parse the output of deSolve and return the variables by name in a list
#' @description This method dispatches on the type of `pars$Xpar`. Adds the variables
#' from the X model to a list and returns it
#' @param deout a [matrix] of outputs from deSolve
#' @param pars a [list] that defines a model
#' @param i the host species index
#' @export
parse_deout_X <- function(deout, pars, i) {
  UseMethod("parse_deout_X", pars$Xpar[[i]])
}
```

#### parse_deout_X.SIS

```{r, eval=F, purl=F}
#' @title Parse the output of deSolve and return variables for the SIS model
#' @description Implements [parse_deout_X] for the SIS model
#' @inheritParams parse_deout_X
#' @return none
#' @export
parse_deout_X.SIS <- function(deout, pars, i) {
  time = deout[,1]
  with(pars$ix$X[[i]],{
    S = deout[,S_ix+1]
    I = deout[,I_ix+1]
    H = S+I
    return(list(time=time, S=S, I=I, H=H))
})}
```


#### parse_deout_X.SIRnew

```{r}
#' @title Parse the output of deSolve and return variables for the SIRnew model
#' @description Implements [parse_deout_X] for the SIRnew model
#' @inheritParams ramp.xde::parse_deout_X
#' @return none
#' @export
parse_deout_X.SIRnew <- function(deout, pars, i) {
  time = deout[,1]
  with(pars$ix$X[[i]],{
    S = deout[,S_ix+1]
    I = deout[,I_ix+1]
    R = deout[,R_ix+1]
    
    H = S+I+R
    return(list(time=time, S=S, I=I, R=R, H=H))
  })}
```

### 14. F_pr  

We require each model to compute the true PR. 

#### Generic 

```{r, eval=F, purl=F}
#' @title Compute the "true" prevalence of infection / parasite rate
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param varslist a list with variables attached by name
#' @param pars a list
#' @param i the host species index
#' @return a [numeric] vector of length `nStrata`
#' @export
F_pr <- function(varslist, pars, i) {
  UseMethod("F_pr", pars$Xpar[[i]])
}
```

#### F_pr.SIS

```{r, eval=F, purl=F}
#' @title Compute the "true" prevalence of infection / parasite rate
#' @description Implements [F_pr] for the SIS model.
#' @inheritParams F_pr
#' @return a [numeric] vector of length `nStrata`
#' @export
F_pr.SIS <- function(varslist, pars, i) {
  pr = with(varslist$XH[[i]], I/H)
  return(pr)
}
```

The function `F_pr` computes the models *true* prevalence using the variable names. It is passed the `varslist` that is created by `parse_deout` (see above). 

#### F_pr.SIRnew

```{r}
#' @title Compute the "true" prevalence of infection / parasite rate
#' @description Implements [F_pr] for the SIRnew model.
#' @inheritParams ramp.xde::F_pr
#' @return a [numeric] vector of length `nStrata`
#' @export
F_pr.SIRnew <- function(varslist, pars,i) {
  pr = with(varslist$XH[[i]], I/H)
  return(pr)
}
```

### 15. HTC  

Similarly, we require each function to return the human transmitting capacity, or `HTC` 

#### Generic 

```{r, eval=F, purl=F}
#' @title Compute the human transmitting capacity
#' @description This method dispatches on the type of `pars$Xpar`.
#' @param pars a [list]
#' @param i the host species index
#' @return none
#' @export
HTC <- function(pars, i) {
  UseMethod("get_inits_X", pars$Xpar[[i]])
}
```

#### HTC.SIS

```{r, eval=F, purl=F}
#' @title Compute the HTC for the SIS model
#' @description Implements [HTC] for the SIS model with demography.
#' @inheritParams ramp.xde::HTC
#' @return a [numeric] vector
#' @export
HTC.SIS <- function(pars, i) {
  with(pars$Xpar[[i]],
    return(c/r)
  )
}
```

#### HTC.SIRnew

```{r}
#' @title Compute the HTC for the SIRnew model
#' @description Implements [HTC] for the SIRnew model with demography.
#' @inheritParams ramp.xde::HTC
#' @return a [numeric] vector
#' @export
HTC.SIRnew <- function(pars, i) {
  with(pars$Xpar[[i]],
       HTC <- c/r,
       return(HTC)
  )
}

```

## Plotting   

### 16. xde_lines_X   

#### xde_lines_X_SIS

```{r, eval=F, purl=F}
#' Add lines for the density of infected individuals for the SIS model
#'
#' @param XH a list with the outputs of parse_deout_X_SIS
#' @param nStrata the number of population strata
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#'
#' @export
xde_lines_X_SIS = function(XH, nStrata, clrs=c("darkblue","darkred"), llty=1){
  with(XH,{
    if(nStrata==1) {
      lines(time, S, col=clrs[1], lty = llty[1])
      lines(time, I, col=clrs[2], lty = llty[1])
    }
    if(nStrata>1){
      if (length(clrs)==2) clrs=matrix(clrs, 2, nStrata)
      if (length(llty)==1) llty=rep(llty, nStrata)

      for(i in 1:nStrata){
        lines(time, S[,i], col=clrs[1,i], lty = llty[i])
        lines(time, I[,i], col=clrs[2,i], lty = llty[i])
      }
    }
  })}
```

#### xde_lines_X_SIRnew

```{r}

#' Add lines for the density of infected individuals for the SIRnew model
#'
#' @param XH a list with the outputs of parse_deout_X_SIRnew
#' @param nStrata the number of population strata
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#'
#' @export
xde_lines_X_SIRnew = function(XH, nStrata, clrs=c("darkblue","darkred","darkgreen"), llty=1){
  with(XH,{
    if(nStrata==1) {
      lines(time, S, col=clrs[1], lty = llty[1])
      lines(time, I, col=clrs[2], lty = llty[1])
      lines(time, R, col=clrs[3], lty = llty[1])
    }
    if(nStrata>1){
      if (length(clrs)==3) clrs=matrix(clrs, 3, nStrata)
      if (length(llty)==1) llty=rep(llty, nStrata)
      
      for(i in 1:nStrata){
        lines(time, S[,i], col=clrs[1,i], lty = llty[i])
        lines(time, I[,i], col=clrs[2,i], lty = llty[i])
        lines(time, R[,i], col=clrs[3,i], lty = llty[i])
      }
    }
  })}

```

### 17. xde_plot_X   

We provide a basic plotting function. The function `xde_lines_X_*` does not have an `S3` version. Instead, it is called by the `S3` function `xde_plot_X`


#### Generic

```{r, eval=F, purl=F}
#' Basic plotting for epidemiological models
#'
#' @param pars a list that defines an `ramp.xde` model (*e.g.*,  generated by `xde_setup()`)
#' @param i the host species index
#' @param clrs a vector of colors
#' @param llty an integer (or integers) to set the `lty` for plotting
#' @param stable a logical: set to FALSE for `orbits` and TRUE for `stable_orbits`
#' @param add_axes a logical: plot axes only if TRUE
#'
#' @export
xde_plot_X = function(pars, i=1, clrs="black", llty=1, stable=FALSE, add_axes=TRUE){
  UseMethod("xde_plot_X", pars$Xpar[[i]])
}
```

#### xde_plot_X.SIS

```{r, eval=F, purl=F}
#' Plot the density of infected individuals for the SIRnew model
#'
#' @inheritParams ramp.xde::xde_plot_X
#' @export
xde_plot_X.SIS = function(pars, i=1, clrs=c("darkblue","darkred"), llty=1, stable=FALSE, add_axes=TRUE){
  vars=with(pars$outputs,if(stable==TRUE){stable_orbits}else{orbits})

  if(add_axes==TRUE)
    with(vars$XH[[i]],
         plot(time, 0*time, type = "n", ylim = c(0, max(H)),
              ylab = "# Infected", xlab = "Time"))


  xde_lines_X_SIS(vars$XH[[i]], pars$Hpar[[i]]$nStrata, clrs, llty)
}
```

#### xde_plot_X.SIRnew

```{r}
#' Plot the density of infected individuals for the SIRnew model
#'
#' @inheritParams ramp.xde::xde_plot_X
#' @export
xde_plot_X.SIRnew = function(pars, i=1, clrs=c("darkblue","darkred","darkgreen"), llty=1, stable=FALSE, add_axes=TRUE){
  vars=with(pars$outputs,if(stable==TRUE){stable_orbits}else{orbits})
  
  if(add_axes==TRUE)
    with(vars$XH[[i]],
         plot(time, 0*time, type = "n", ylim = c(0, max(H)),
              ylab = "No of. Infected", xlab = "Time"))
  
  
  xde_lines_X_SIRnew(vars$XH[[i]], pars$Hpar[[i]]$nStrata, clrs, llty)
}
```


