---
title: "SIR-xde (Susceptible-Infected-Recovered) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIR-xde (Susceptible-Infected-Recovered) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIR (Susceptible-Infected-Recovered) human xde model model fulfills the generic interface of the human population component. It is the simplest model of endemic diseases in humans with recovery.

# 

We subdivide a population into susceptible ($S$)  infected and infectious ($I$) and recovered ($R$) individuals, where the total population is $H = S+I+R.$ We assume the force of infection ($h$, FoI) is linearly proportional to the EIR: $h = b \times EIR.$ In its general form, with births ($B(H)$) and deaths (at the per-capita rate $\mu$), the generalized SIR_xde dynamics are:

$$
\begin{array}{rl}
\dot{S} &= -h S + B(H) -\mu S\\
\dot{I} &= h S  - rI - \mu I\\
\dot{R} &= rI-\mu R
\end{array}
$$
We are assuming that a population could be stratified, such that the variables and parameter are all vectors with length `nStrata.`

# Equilibrium Solutions

This model would have no interesting dynamics at equilibrium as the total population $H$ would tend to the recovered population $R$.

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(ramp.library)
library(deSolve)
library(data.table)
library(ggplot2)
```


```{r, echo=FALSE} 
#devtools::load_all()
```
## Using Setup
Here we use `ramp.xds::xds_setup` to set up the model object. We solve the model to equilibrium and check with the analytic steady state using the  `ramp.xds::xde_steady_state` and compare the results.



```{r}
test_SIR <- xds_setup(MYZname ="macdonald", Xname="SIR")
```

```{r}
xds_solve(test_SIR, 365*10)-> test_SIR
unlist(list_Xvars(test_SIR$outputs$last_y, test_SIR, 1)) -> out1
```

## Using the   `ramp.xds::xde_steady_state`

```{r}
foi_eq = test_SIR$Xpar[[1]]$b*tail(test_SIR$outputs$terms$EIR,1)
xde_steady_state_X(1/365, 1000, test_SIR$Xpar[[1]]) -> out2
```
Then we compare the analytic and numerical solution
```{r}
sum(abs(out2 - out1))

```

