---
title: "Compartmental Models"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette 
vignette: >
  %\VignetteIndexEntry{Compartmental Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, message=FALSE, warning=FALSE}
library(knitr)
library(ramp.xds)
library(ramp.library)
library(deSolve)
library(ramp.library)
```

```{r, echo=F}
#devtools::load_all()
```

In this vignette, we describe four basic compartmental models for mosquito-borne pathogens commonly used for humans in `ramp.library`. The models are named by their states: susceptible (S) - exposed (E) infected (I) - recovered and immune (R), and vaccinated (V). According to their setup, not all models include all of the states. There are four compartmental models: SIR, SIRS, SEIR, and SEIRV. The other models including SIS, SEI and SIP and SEIS are described in the `ramp.xds`.
  
##  Model Variables

+ $S$ is the density of susceptible humans

+ $E$ is the density of exposed humans

+ $I$ is the density of infected humans

+ $R$ is the density of recovered and immune humans

+ $V$ is the density of vaccinated humans

+ $H$ is the density of humans

## Parameters

+ $r$ is the rate infections clear

+ $b$ is the fraction of bites by infective mosquitoes that transmit parasites and cause an infection. 

+ $c$ is the fraction of bites on an infectious human that would infect a mosquito.

+ $\varepsilon$ proportion of recovered individuals progressing into vaccinated class

+ $\alpha$ is a proportion of vaccinated humans

+ $\tau$ is the rate at which exposed humans become infectious (incubation rate)

+ $\gamma$ rate at which recovered human loss their immunity

+ $\delta$ proportion of recovered humans progressing to vaccinated class

## Dynamics

The models defined herein is defined in two parts. To model exposure and infection (*i.e.* the conversion of EIR into FoI (h)), The equations are formulated around $h$. Under the default model, we get the relationship $h=bE$, where E is the daily EIR:


### The dynamics of Susceptible-Infected- Recovered (SIR) are given by: 

$$
\begin{eqnarray*}
\frac{dS}{dt} &=&B(H) - h S - \mu S\\
\frac{dI}{dt} &=& h S- (r +\mu) I\\
\frac{dR}{dt} &=& r I- \mu R
\end{eqnarray*}
$$
Without demography i.e $ B(H) = \mu = 0$, SIR model has steady states as $$ 
\bar{S} = 0, \quad  \bar{I} =0, \quad \bar{R} = H$$.

### The dynamics of Susceptible-Infected- Recovered -Suscepitible  (SIRS) are given by:

$$
\begin{eqnarray*}
\frac{dS}{dt} &=& B(H) - h S + \gamma R - \mu S\\
\frac{dI}{dt} &=& h S- (r +\mu) I\\
\frac{dR}{dt} &=& r I- (\mu+ \gamma) R
\end{eqnarray*}
$$
Without demography, the SIRS model has the following steady states as
 
$$
\begin{eqnarray*}
\bar{I} &=& \frac{h H\gamma}{r\gamma + h(\gamma+r) }\\
\bar{R} &=& \frac{h r H}{r\gamma + h(\gamma+r) }\\
\bar{S} &=& H- \bar{I}-\bar{R}
\end{eqnarray*}
$$


### The dynamics of Susceptible -Exposed -Infected- Recovered  (SEIR) are given by: 

$$
\begin{eqnarray*}
\frac{dS}{dt} &=& B(H) - h S - \mu S\\
\frac{dE}{dt} &=& hS - \tau E- \mu E\\
\frac{dI}{dt} &=& \tau E- (r +\mu) I\\
\frac{dR}{dt} &=& r I- \mu R
\end{eqnarray*}
$$
Without demography, the SEIR model has the steady states given as 
$$ \bar{S} = 0, \quad \bar{E} = 0,\quad \bar{I} = 0,\quad \bar{R} = H$$

### The dynamics of Susceptible-Infected- Recovered -Suscepitible- Vaccinated (SEIRV)  are based on the  are given by: 

$$\begin{eqnarray}
\frac{dS}{dt} &=&(1-\alpha)B(H) - h S + \gamma R - \mu S\\
\frac{dE}{dt} &=& h S - (\tau  + \mu)E\\
\frac{dI}{dt} &= &\tau E - (r +\mu) I\\
\frac{dR}{dt} &=& (1-\varepsilon) r I- (\gamma+\mu) R\\
\frac{dV}{dt} &=& \alpha B(H)+ \varepsilon r I -\mu V
\end{eqnarray}
$$
Without demography, the SEIRV has the following steady states:
$$ \bar{S} = 0, \quad \bar{E} = 0,\quad \bar{I} = 0,\quad \bar{R} = 0,\quad  \bar{V}= H$$


## Terms 

**Net Infectiousness**

True prevalence is:

$$x = \frac{X}{H}.$$ 

In our implementation, net infectiousness (NI) is linearly proportional to prevalence:

$$ni = c x.$$

**Human Transmitting Capacity**

After exposure, a human would remain infected for $1/r$ days, transmitting with probability $c$ so:

$$HTC= c/r$$

## Example `ramp.xds` Setup

We run each of the models using a default setup with 1- stratum  to equilibrium and compare our results with the analytic steady states given above

```{r}
library(ramp.xds)
library(ramp.library)
library(deSolve)
library(data.table)
library(ggplot2)
```

```{r}
#devtools::load_all()
```

```{r}
test_SIR<- xds_setup(MYname="macdonald", Xname="SIR")
xds_solve(test_SIR, 365*5) -> test_SIR
foi_eq = tail(test_SIR$outputs$orbits$XH[[1]]$foi,1)
unlist(get_XH_vars(test_SIR$outputs$last_y, test_SIR, 1)) -> inf 
steady_state_X(foi_eq, 1000, test_SIR, 1) -> ss
xds_plot_X(test_SIR)
```

```{r}
test_SIRS<- xds_setup(MYname="macdonald", Xname="SIRS")
xds_solve(test_SIRS, 365*3) -> test_SIRS
foi_eq = tail(test_SIRS$outputs$orbits$XH[[1]]$foi,1)
unlist(get_XH_vars(test_SIRS$outputs$last_y, test_SIRS, 1)) -> inf_SIRS
steady_state_X(foi_eq, 1000, test_SIRS, 1) -> ss_SIRS
xds_plot_X(test_SIRS)
```

```{r}
test_SEIR<- xds_setup(MYname="macdonald", Xname="SEIR")
xds_solve(test_SEIR, 365*5) -> test_SEIR
foi_eq = tail(test_SEIR$outputs$orbits$XH[[1]]$foi,1)
unlist(get_XH_vars(test_SEIR$outputs$last_y, test_SEIR, 1)) -> inf_SEIR
steady_state_X(foi_eq, 1000, test_SEIR, 1) -> ss_SEIR
xds_plot_X(test_SEIR)
```

```{r}
test_SEIRV<- xds_setup(MYname="macdonald", Xname="SEIRV")
xds_solve(test_SEIRV, 365*5) -> test_SEIRV
foi_eq = tail(test_SEIRV$outputs$orbits$XH[[1]]$foi,1)
unlist(get_XH_vars(test_SEIRV$outputs$last_y, test_SEIRV, 1)) -> inf_SEIRV
steady_state_X(foi_eq, 1000, test_SEIRV, 1) -> ss_SEIRV
xds_plot_X(test_SEIRV)
```



## References

1. Ross R. Report on the Prevention of Malaria in Mauritius. London: Waterlow; 1908.
2. Ross R. The Prevention of Malaria. 2nd ed. London: John Murray; 1911.
3. Smith DL, Battle KE, Hay SI, Barker CM, Scott TW, McKenzie FE. Ross, Macdonald, and a theory for the dynamics and control of mosquito-transmitted pathogens. PLoS Pathog. 2012;8: e1002588. doi:10.1371/journal.ppat.1002588


